Parametricity Consequences of GADTs Representations

### 2021-02-16 20-gadts/GADT-story.tex
setting up Par section

Just as for ADTs, we can represent GADTs as either Church encodings or
primitive data types. But whereas the two representations are always
isomorphic for ADTs, they not, in general, for GADTs. We will see this
in Section~\ref{??} below. {\color{blue} But we get
  rearrange-transform property for primitive rep since this is a
  consequence of functoriality.}

{\color{blue}
Forced to choose, a programmer would likely find the idea that a GADT
contains data that are not specified by its syntax disturbing. What,
they might ask, should a data type contain other than data that are
constructed using its data constuctors? Why should a GADT contain
``secret'' elements that are only attainable by mapping? On the other
hand, a semanticist would likely find the primitive representation of
GADTs more reasonable. After all, they would likely find the
nonfunctorial nature of the Church encoding unnerving at best. After
all, they would argue, the data in a GADT shouldn't disappear just
because you map a function over it! The fact that this happens for
Church encodings of GADTs --- i.e., for GADTs as specified by their
syntax --- highlights how GADTs {\em do not} generalize the essential
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense are GADTs data types at all?}

{\color{blue} THIS TRANSITION FROM EQUIVALENCE OF REPRESENTATION TO
  PARAMETRICITY GOES SOMEWHERE:
Non-uniformity renders build, therefore short cut, for GADTs
impossible. Can't even formulate! It is possible there could be
another way to prove equivalence, but we know the two representations
are not equivalent because we have a counterexample.

The fact that GADTs don't validate any such iso has important
consequences for their parametricity properties.}

@@

{\color{blue} Say somewhere: A relation is a triple $(A,B,R)$, where
  $R$ is a relation between sets $A$ and $B$.  A function from
  $(A,B,R)$ to $(A',B',R')$ is a pair $(f : A \to A',g : B \to B')$ of
  functions on sets such that $(f a,g\,b) \in R'$ if $(a,b) \in R$. We
  may write $R : \rel(A,B)$ for $(A,B,R)$.  If $R : \rel(A,B)$ we
  write $\pi_1 R$ and $\pi_2 R$ for the {\em domain} $A$ of $R$ and
  the {\em codomain} $B$ of $R$, respectively, and assume $\pi_1$ and
  $\pi_2$ are surjective. We write $\Eq_A = (A,A,\{(x,x)~|~ x \in
  A\})$ for the {\em equality relation} on the set $A$.}

>>>>

Just as for ADTs, we can represent GADTs as either Church encodings or
primitive data types. But whereas the two representations are always
isomorphic for ADTs, they not, in general, for GADTs. We will see this
in the next section.

\section{Non-Equivalence of Church Encodings and Primitive
  Representations of GADTs}\label{sec:non-equiv}

Intuitively it seems the two representations might be different. One
is functorial, one is not. Non-uniform nature of GADTs renders build,
therefore short cut, for GADTs impossible. Can't even formulate! Of
course, there could be another way to prove equivalence.  But, in
fact, CEs and primitive representations are not equivalent for
arbitrary GADTs, as the following example shows.

\begin{example}
Use our favorite GADT to establish non-equivalence of the two reps
\end{example}

Forced to choose, a programmer would likely find the idea that a GADT
contains data that are not specified by its syntax disturbing. What,
they might ask, should a data type contain other than data that are
constructed using its data constuctors? Why should a GADT contain
``secret'' elements that are only attainable by mapping? On the other
hand, a semanticist would likely find the primitive representation of
GADTs more reasonable. After all, they would likely find the
nonfunctorial nature of the Church encoding unnerving at best. After
all, they would argue, the data in a GADT shouldn't disappear just
because you map a function over it! The fact that this happens for
Church encodings of GADTs --- i.e., for GADTs as specified by their
syntax --- highlights how GADTs {\em do not} generalize the essential
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense are GADTs data types at all?

The disappointment that GADTs don't validate any such equivalence is
further compounded by the fact that the the two representations have
very different parametricity properties.  {\color{blue} For example,
  we get rearrange-transform property for primitive rep since this is
  a consequence of functoriality, but don't get it for Church
  Encodings.} We explore the differences in parametricity results for
the two representations in the next section.

@@

{\color{blue} Say somewhere: A relation is a triple $(A,B,R)$, where
  $R$ is a relation between sets $A$ and $B$.  A function from
  $(A,B,R)$ to $(A',B',R')$ is a pair $(f : A \to A',g : B \to B')$ of
  functions on sets such that $(f a,g\,b) \in R'$ if $(a,b) \in R$. We
  may write $R : \rel(A,B)$ for $(A,B,R)$.  If $R : \rel(A,B)$ we
  write $\pi_1 R$ and $\pi_2 R$ for the {\em domain} $A$ of $R$ and
  the {\em codomain} $B$ of $R$, respectively, and assume $\pi_1$ and
  $\pi_2$ are surjective. We write $\Eq_A = (A,A,\{(x,x)~|~ x \in
  A\})$ for the {\em equality relation} on the set $A$. ALSO, WANT OUR
RELATIONS TO BE ONTO.}
